\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-11-16 review_response.Rnw

\Bigtxt{Reviewer 1:}

*Main text*

I didn't find it obvious from the text whether the setstep and comb functions would handle missing data with interpolation or NA. (It's NA, as it turns out.) Perhaps it is because I don't do a lot of work in this domain, though.

{\it This was clarified in the revision, page 6 paragraph one:

`Both functions treat missing data as NA values, either for observations that exceed the allowable tolerance for the differ argument of setstep or for time series that do not overlap given the method argument passed to comb.' 
}

@@

*Package performance*

Some of the functions can spend a surprising amount of time parsing. For example, the call: all\_params\_dtrng(`hudscwq', dtrng = c(`09/01/2013', `10/01/2013')) took 19 seconds on my machine, and running under a profiler showed that about 12-3 of those seconds were spent parsing the data, not downloading it. The SWMPr::parser code calls htmlTreeParse, which is where all that time is being spent; I found that changing it to xmlTreeParse made almost all of that time go away. The only difference I noticed was that htmlTreeParse converts all of the tag names to lowercase, while xmlTreeParse does not. This can be easily fixed by having the parser() function change the `out' data frame's names to lowercase before returning. With these two modifications, the elapsed time drops to 6-7 seconds--the bulk of the time now being spent downloading the data (as far as I can tell).

{\it That is a fantastic suggestion and I've made these changes in the development repository of the package.  I noticed an immediate increase in performance with the import functions.  I will push the new version to CRAN if or when the paper is accepted for publication. 
}

@@

*shiny\_comp app*

The Shiny app that uses Leaflet is using a deprecated API. The new API is documented at https://rstudio.github.io/leaflet/ and I think it would significantly simplify the code. I can lend a hand if the author wants.

{\it I would be interested in working together to improve the code.  I made the app by borrowing an example from the rstudio GitHub repo (https://github.com/rstudio/leaflet/tree/master/inst/legacy/examples/population) and tweaking the code for my application.  I admit that I have a fuzzy understanding of using leaflets maps reactively and I have always had trouble editing the code as a result.  Simplification with the new API would be great not only for performance, but also to help me better understand reactive programming.}

@@

The Leaflet map could use a color legend, which is possible with the new Leaflet API. (It could use a radius legend too, but that is not yet a feature of the Leaflet package.)

There is a feature in the new Leaflet API that lets you have a label http://leaflet.github.io/Leaflet.label/ for each circle that is either always shown, or shown on mouseover. This is a nicer experience for the user than having to click. This feature has not yet made it to CRAN, but it is available in the GitHub master branch of Leaflet.

@@

It would be interesting if the Leaflet app could let you select multiple stations and see the data plotted together.

{\it Again, these are all great suggestions but I would probably need assistance to implement these changes with the new API.}

@@

*shiny\_summary app*

In the file https://github.com/fawda123/swmp\_summary/blob/master/server.R, the plotInput should not be a reactive(), but a regular function that takes no arguments. This is because reactives cache their values, and should not be executed for their side effects (in this case, plotting). As the app is written now, there doesn't appear to be an actual user-visible bug, but a change as simple as setting the plot width to ``100\%'' would introduce a bug (I'd expect it to fail to redraw the plot when the window was resized).

{\it Yes, the plot fails to render on shinyapps when the window is resized.  I had not noticed the bug so I appreciate the comment.  I've changed the code in server.R to make plotInput a regular function.  I've redeployed the app and the bug has been fixed.}

@@

``SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.  For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.  The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for the time series (Helsel, 2012).  A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages (e.g., cents, McLeod et al., 2014).  However, the functions in SWMPr can be used to identify censored data based on the appropriate QAQC flag for a given parameter.  Viewing this information can be helpful for determining how to further process the data with the qaqc function or alternative methods outside of SWMPr.  The qaqcchk function returns a data.frame of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.  SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.''
}

Second, in the ``Applications using the SWMPr package'' section near the end of the manuscript, simple linear regression (SLR) is identified as a technique used to summarize trends in a parameter over time. This particular statistical method may or may not be appropriate for a given data set. Alternative trend analysis methods exist, including several that treat censored values appropriately (again, Helsel 2012 is a useful reference). It is unclear from the manuscript what alternatives to SLR may be available in SWMPr (e.g., maximum likelihood estimation, Akritas-Theil-Sen nonparametric regression, others). Therefore, clarifying text that either identifies the availability of other trend analysis procedures in SWMPr or provides caveats to the use of SLR (e.g., used only for initial exploratory analysis) should be added to this section.

@@

Page 12, paragraph one: `More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for exploratory purposes only.'

Added to app: `Please note that the use of simple regression to identify trends is for exploratory purposes only and may not be appropriate for all datasets.'
}

Reference:
Helsel, D.R. 2012. Statistics for Censored Environmental Data Using Minitab and R. 2nd edition, John Wiley \& Sons, Inc. Hoboken, New Jersey. 324pp.

{\it This reference was added for the text additions above.}

### 2015-11-16 swmpr_manu.tex
response to first reviewer done, need to proofread, recompile

@@

% table for retrieval functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Retrieval functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?all\_params} for individual functions or \code{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \code{exportAllParamsXMLNew} function on web services.\tabularnewline
\code{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \code{exportAllParamsDateRangeXMLNew}.\tabularnewline
\code{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
\code{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \code{exportSingleParamXMLNew} function on web services.\tabularnewline
\code{site\_codes}&Get metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
\code{site\_codes\_ind}&Get metadata for all stations at a single site, wrapper to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

\end{table}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import all paramaters for the station
# three most recent records
exdat <- all_params('apadbwq', Max = 3, trace = F)
exdat
##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
## 1 2015-11-03 11:15:00   26      0     45        0  29     0     78
## 2 2015-11-03 11:30:00   26      0     46        0  30     0     76
## 3 2015-11-03 11:45:00   26      0     46        0  30     0     75
##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
## 1        0      5        0     2       0  8    0    2      0       NA
## 2        0      5        0     2       0  8    0    5      0       NA
## 3        0      5        0     2       0  8    0    5      0       NA
##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
## 1         -2    NA      -1      2     NA        3         
## 2         -2    NA      -1      2     NA        3         
## 3         -2    NA      -1      2     NA        3
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

% attributes table
%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\code{\"swmpr\"} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Attributes of a \code{"swmpr"} object that describe characteristics of the data.\label{tab:attributes}} 
\begin{center}
\begin{tabular}{lp{0.75in}p{3.25in}}
\hline\hline
\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{names}&character&Column names of the entire data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{row.names}&integer&Row names of the data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{"data.frame"}.\tabularnewline
\code{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{'do\_mgl'}, \code{'turb'}, etc.\tabularnewline
\code{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
\code{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
\code{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually \code{"POSIXct"} unless data have been aggregated.\tabularnewline
\hline
\end{tabular}\end{center}

\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \code{"swmpr"} objects and may not be cities in the United States.  For example, \code{America/Jamaica} is used for Eastern Standard Time.\end{table}

@@

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?comb} for individual functions or \\code{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Organizing functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?comb} for individual functions or \code{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.\tabularnewline
\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.\tabularnewline
\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.\tabularnewline
\code{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.\tabularnewline
\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function in the base installation.\tabularnewline
\hline
\end{tabular}\end{center}

\end{table}

@@

The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.  The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.  Data are combined by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.  The time series for merging depends on the value passed to the \code{method} argument.  Passing \code{'union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects, whereas \code{'intersect'} will create a continuous time series from the set of dates that are shared between input objects.  A character string or numeric vector can also be used to specify which of the input objects to use as the master time series for combining.  As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.

@@

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the \pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.  This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \citep{Odum56}.\tabularnewline
\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \code{decompose} \citep{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\code{hist}&Plot a histogram for a single variable.\tabularnewline
\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object.\tabularnewline
\code{overplot}&Plot multiple time series on the same y-axis.\tabularnewline
\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.\tabularnewline
\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a single parameter.\tabularnewline
\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average, passed to \code{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

\end{table}

>>>>

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in loadNamespace(j <- i[[1L]], c(lib.loc, .libPaths()), versionCheck = vI[[j]]): there is no package called 'Rcpp'}}\end{kframe}
\end{knitrout}

@@

% table for retrieval functions
\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import all paramaters for the station
# three most recent records
exdat <- all_params('apadbwq', Max = 3, trace = F)
\end{verbatim}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}all\_params"{}}}\begin{verbatim}
exdat
\end{verbatim}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'exdat' not found}}\end{kframe}
\end{knitrout}

@@

% attributes table
\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}

@@

% table for organize functions
\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}

@@

The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.  The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.  Data are combined by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.  The time series for merging depends on the value passed to the \code{method} argument.  Passing \code{'union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects, whereas \code{'intersect'} will create a continuous time series from the set of dates that are shared between input objects.  A character string or numeric vector can also be used to specify which of the input objects to use as the master time series for combining.  As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.  Both functions treat missing data as \code{NA} values, either for observations that exceed the allowable tolerance for the \code{differ} argument of \code{setstep} or for time series that do not overlap given the \code{method} argument passed to \code{comb}.

@@

% table for analysis functions
\begin{kframe}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}
